# Appendix: Data cleaning {.unnumbered}

```{r}
# source("_load.r")

```

## Cleaning latitude and longitude

Several of the VFWD sites appear to have latitude or longitude miscoded. I determined that by looking up their locations on a map (with Google Maps) using their stated lat-long values and comparing these to the location names in the data file. Based on this, I concluded that:

-   Site id 502377, White Creek Rt 153 bridge, with stated lat, long of (42.2363891, -73.2436065) should have latitude start with 43, not 42.

-   Site id 502378, White Creek ups Rt 153, with stated lat, long of (42.2356796, -73.2388) should have latitude start with 43, not 42.

-   Site id 502397, Bourn Book Little Mud Pond, with stated lat, long of (43.1402816, -73.9894409) should have longitude start with -72 rather than -73.

-   Site id 521549, White Creek off Kent Hollow Rd, with stated lat, long of (42.2342491, -73.2149276 ) should have latitude start with 43, not 42.

The table below shows the sites for which I changed latitude or longitude values (the variable names ending in "\_raw" show the data as reported; the other variables are values as adjusted).

```{r}
#| include: true

bkill %>%
  filter(latitude != latitude_raw | longitude != longitude_raw) %>%
  select(src, siteid, locdesc, latitude_raw, longitude_raw, latitude, longitude) %>%
  distinct() %>%
  arrange(siteid) %>%
  kbl() %>%
  kable_styling()

```

## Determining name of stream

The NYSDEC data have a variable (`LOCATION`) that identifies each river or stream uniquely but the VFWD data do not.

It is important to have a unique identifier for each stream because the "river miles" variable measures the number of miles from the mouth of a stream to a given site, and it is meaningless without knowing which stream it refers to.

I created a unique short identifier and a unique long identifier for each stream as follows:

-   NYSDEC: I used the identifier on the file for the short name, and defined a long name based on the location description

-   VFDW: I parsed the location description to identify each stream uniquely, and defined short and long identifiers based on the location description

I list the identifiers below.

```{r}
#| include: true
bkill %>%
  select(streamid, stream) %>%
  distinct() %>%
  arrange(streamid) %>%
  kbl() %>%
  kable_styling()
```

## Adding unique site ids

I created a unique site id for each location:

-   NYS DEC sites: "NYSDEC\_" followed by BAS_LOC_RM value

-   VFWD sites: "VFWD\_" followed by the SITE ID value

## Adding state abbreviations

After adjusting latitude and longitude, all observations from the VFWD file are in Vermont. However, 4 of the 58 observations in the NYSDEC file are in Vermont. All 4 are on the Battenkill, just over the border. As a result, I added a state abbreviation to the file so that we know which state the observations are in.

Here are the 4 observations from the NYSDEC file that are in Vermont:

```{r}
#| include: true

bkill %>%
  filter(src=="NYSDEC", stabbr=="VT") %>%
  select(src, stabbr, streamid, stream, locdesc, latitude, longitude) %>%
  arrange(longitude, latitude) %>%
  kbl(format.args = list(decimal.mark = '.', digits=4)) %>%
  kable_styling()

```

## Cleaning the Hilsenhoff Biotic Index (HBI)

The HBI is supposed to range from 0 to 10. All of the observations in the data were in this range except for one.

The observation at site id 502390, Lye Brook Confluence, on 1990-09-28 was 26.5. Other observations near Lye Brook in the same time period (1990-1995) ranged between 1.44 and 1.92. Based on this, I assumed this was a typo, probably with the decimal in the wrong place, so I divided it by 10 giving 2.65. However, I flagged the observation as potentially bad, so that we can drop it instead of altering, if desired.

## Adjusted data file

I created an adjusted data file with all of the changes above and used it for all of the analysis in this document.
