# Ephemeroptera, Plecoptera, and Trichoptera (the EPT Index)

The EPT Index is named for three orders of aquatic insects that are common in the benthic (bottom dwelling) macroinvertebrate community:

-   Ephemeroptera (mayflies)

-   Plecoptera (stoneflies), and

-   Trichoptera (caddisflies).

The EPT Index is based on the premise that high-quality streams usually have the greatest species richness. Many aquatic insect species are intolerant of pollutants and will not be found in polluted waters. **The greater the pollution, the lower the species richness expected, as only a few species are pollutant tolerant.** [@mcquaidEPTIndex2012]

```{r startup}
source(here::here("r", "libraries.r"))
# bkill <- readRDS(here::here("data", "bkill.rds"))
bkill <- readRDS(here::here("data", "gis", "bkill.rds"))
```


## EPT richnesss measure over time at sites with repeated observations

```{r}
#| include: false
#| fig-height: 10
#| fig-width: 7.5

repeats <- bkill %>%
  filter(!is.na(ept_rich)) %>%
  count(locid, locdesc, sort = TRUE) %>%
  filter(n > 1)

p <- bkill %>%
  filter(locid %in% (repeats %>% filter(n >=4) %>% pull(locid))) %>%
  ggplot(aes(date, ept_rich)) +
  geom_point(size=2.25, colour="blue") +
  # geom_smooth(se=FALSE) +
  scale_y_continuous(limits=c(0, NA)) +
  facet_wrap(~locid+locdesc, ncol=3) +
  labs(x=NULL) +
  ggtitle("EPT richness measure over time at locations that had at least 4 repeated observations",
          subtitle="Higher values mean greater richness") +
  theme_bw()
p
ggsave(here::here("results", "eptrich_over_time.png"), height=10, width=7.5, units="in", scale = 1)


```



```{r}
#| include: true
#| fig-height: 10
#| fig-width: 8.5

repeats <- bkill %>%
  filter(!is.na(ept_rich)) %>%
  count(locid, locdesc, sort = TRUE) %>%
  filter(n > 1)

p <- bkill %>%
  filter(locid %in% (repeats %>% filter(n >=3) %>% pull(locid))) %>%
  ggplot(aes(date, ept_rich)) +
  geom_point(size=2.25, colour="blue") +
  # geom_smooth(se=FALSE) +
  scale_y_continuous(limits=c(0, NA)) +
  facet_wrap(~locid+locdesc, ncol=4) +
  labs(x=NULL) +
  ggtitle("EPT richness measure over time at locations that had at least 3 repeated observations",
          subtitle="Higher values mean greater richness") +
  theme_bw()
p
ggsave(here::here("results", "eptrich_over_time3.png"), height=10, width=8.5, units="in", scale = 1)


```